---
title: "Data Visualization and Exploratory Data Analysis"
author: "Jessica Lavery"
date: "9/26/2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# changing the size of a figure
# out.width/out.height control the size of the figure inserted into your document
knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

library(tidyverse)
library(ggridges)
```

# Read in data from rnoaa package

```{r, message = FALSE}
weather_df = 
  rnoaa::meteo_pull_monitors(c("USW00094728", "USC00519397", "USS0023B17S"),
                      var = c("PRCP", "TMIN", "TMAX"), 
                      date_min = "2017-01-01",
                      date_max = "2017-12-31") %>%
  mutate(
    name = recode(id, USW00094728 = "CentralPark_NY", 
                      USC00519397 = "Waikiki_HA",
                      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) %>%
  select(name, id, everything())

weather_df
```

# ggplot

## Basic scatterplot

```{r}
# via ggplot w/o pipe
ggplot(data = weather_df, aes(x = tmin, y = tmax)) +
  geom_point()

# alternatively
weather_df %>%
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point()

# save plot as object
plot_weather = 
  weather_df %>%
  ggplot(aes(x = tmin, y = tmax)) 

# useful to create initial plot and then edit or to create a plot and then save and export
plot_weather + geom_point()

class(plot_weather) # gg/ggplot
```

## Add colors

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = 0.4) + #use alpha to add transparency
  geom_smooth(se = FALSE)

# vs including color=name on the ggplot call as opposed to the geom_point
# color then applies to each of the geometries as opposed to just one
ggplot(weather_df, aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = 0.4) + #use alpha to add transparency
  geom_smooth(se = FALSE)
```

## Panel (facet) by weather stations

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .5) +
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name) #want name to be the columns
```

## Look at data by date, add bubble 

```{r}
ggplot(weather_df, aes(x = date, y = tmax, color = name)) + 
  geom_point(aes(size = prcp), alpha = .5) +
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name)
```

```{r}
ggplot(weather_df, aes(x = date, y = tmax, color = name)) + 
  geom_point(aes(size = prcp), alpha = .5) +
  geom_smooth(se = FALSE)
```

## 2-d density

Useful to show a lot of data points that my overlap. 

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_hex()
```


## Learning assessment 
```{r}
weather_df %>% 
  filter(name == "CentralPark_NY") %>% 
  mutate(tmin_F = weathermetrics::celsius.to.fahrenheit(tmin),
         tmax_F = weathermetrics::celsius.to.fahrenheit(tmax)) %>% 
  ggplot(aes(x = tmin_F, y = tmax_F)) + 
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE)
```

Compare the results from these two plots:
```{r}
#sets the color to blue
ggplot(weather_df) + geom_point(aes(x = tmax, y = tmin), color = "blue")

#tells the aesthetic that "blue" indicates the groups that should be used to create colors
ggplot(weather_df) + geom_point(aes(x = tmax, y = tmin, color = "blue"))
```

# Univariate plots

## Histogram
```{r}
ggplot(weather_df, aes(x = tmax, color = name, fill = name)) + 
  geom_histogram(position = "dodge", binwidth = 2)
```

## Density plot
```{r}
#adjust ranges from 0 to 1, 0 is most granular, 1 is most smooth
ggplot(weather_df, aes(x = tmax, fill = name)) + 
  geom_density(alpha = .4, adjust = .9, color = "orange") 
```

## Boxplot
```{r}
ggplot(weather_df, aes(x = name, y = tmax)) + 
  geom_boxplot()
```

## Violin plot
```{r}
ggplot(weather_df, aes(x = name, y = tmax)) + 
  geom_violin(aes(fill = name), color = "blue", alpha = .5) + 
  stat_summary(fun.y = median, geom = "point", color = "blue", size = 4)
```

## Ridge plot
```{r}
ggplot(weather_df, aes(x = tmax, y = name)) + 
  geom_density_ridges(scale = .85)
```

## Learning Assessment 
Make plots that compare precipitation across locations. Try a histogram, a density plot, a boxplot, a violin plot, and a ridgeplot; use aesthetic mappings to make your figure readable.

```{r}
base <- ggplot(data = weather_df, aes(x = prcp, color = name))

# histogram
base + 
  geom_histogram(aes(fill = name))

# density
base +
  geom_density()

# boxplot
base + 
  geom_boxplot(aes(x = name, y = prcp))

# violin plot
base + 
  geom_violin(aes(x = name, y = prcp))

# ridge plot
ggplot(weather_df, aes(x = tmax, y = name)) + 
  geom_density_ridges(scale = .85)
```

# Saving plots
```{r}
weather_plot = ggplot(weather_df, aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) 

## bad practice/not reproducible to hit export -> save as PDF
ggsave("weather_plot.pdf", weather_plot, width = 8, height = 5)
```
