---
title: "Exploratory Data Analysis"
author: "Jessica Lavery"
date: "9/26/2019"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(viridis)
library(knitr)

knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      fig.width = 8,
                      fig.height = 6,
                      out.width = "90%")

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

```{r, cache = TRUE}
weather_df = 
  rnoaa::meteo_pull_monitors(c("USW00094728", "USC00519397", "USS0023B17S"),
                      var = c("PRCP", "TMIN", "TMAX"), 
                      date_min = "2017-01-01",
                      date_max = "2017-12-31") %>%
  mutate(
    name = recode(id, USW00094728 = "CentralPark_NY", 
                      USC00519397 = "Waikiki_HA",
                      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
    month = lubridate::floor_date(date, unit = "month")) %>%
  select(name, id, date, month, everything())
```

## `group_by` and counting
```{r}
weather_df %>% 
  group_by(name, month)
```

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarize(n_obs = n())
```

### `n_distinct`

```{r}
weather_df %>% 
  group_by(month) %>% 
  summarize(n_obs = n(),
            n_unique = n_distinct(date))
```


### `count`

Count shortens using `group_by` and `summarize`.

```{r}
# default varible name for count is n
weather_df %>% 
  count(name)

# can also customize this
weather_df %>% 
  count(name, name = "n_days")
```

** Never use base R's table() function ** 

```{r}
tbl <- weather_df %>% 
  pull(name) %>% 
  table()

# print the result
tbl

# result is of class "table", not a data frame! 
class(tbl)

# same as table(weather_df$name)
```

## Kable

Let's make a nice table

```{r}
weather_df %>% 
  count(name) %>% 
  knitr::kable()
```

## 2x2 tables

```{r}
weather_df %>% 
  filter(name != "Waikiki_HA") %>% 
  mutate(
    cold = case_when(
      tmax < 5  ~ "cold",
      tmax >= 5 ~ "not cold",
      TRUE      ~ ""
    )
  ) %>% 
  group_by(name, cold) %>% 
  count() %>% 
  pivot_wider(
    names_from = cold,
    values_from = n
  )
```

Alternative approach using `janitor::tabyl()`

https://cran.r-project.org/web/packages/janitor/vignettes/tabyls.html

```{r}
weather_df %>% 
  filter(name != "Waikiki_HA") %>% 
  mutate(
    cold = case_when(
      tmax < 5  ~ "cold",
      tmax >= 5 ~ "not cold",
      TRUE      ~ ""
    )
  ) %>% 
  janitor::tabyl(name, cold) %>%
  janitor::adorn_percentages("row") %>%
  janitor::adorn_pct_formatting(digits = 2) %>%
  janitor::adorn_ns() %>% 
  kable()
```

## general summaries

By deafult, any time that you compute a numeric summary of something that includes an NA, R returns an NA. Options: drop these rows from the data, modify function to ignore NAs in computation via na.rm = TRUE. Don't do the latter option by default, look at the missing values and make sure you understand what's going on. 

Can also pipe in the summary into ggplot to create a visual of the summary that was created. 

```{r}
weather_df %>% 
  group_by(name, month) %>%
  summarize(
    n = n(),
    mean_tmax = mean(tmax, na.rm = TRUE),
    sd_tmax = sd(tmax, na.rm = TRUE),
    median_prcp = median(prcp, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = month, y = mean_tmax, color = name)) + 
    geom_point() + geom_line() + 
    theme(legend.position = "bottom")
```

### user-friendly summaries

The output won't be in "tidy" format, but will be more user-friendly. 

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE)
  ) %>% 
  pivot_wider(
    names_from = name,
    values_from = mean_tmax) %>% 
  knitr::kable()
```

### ungrouping

```{r}
weather_df %>% 
  group_by(name) %>% 
  ungroup()
```

### grouping & mutating

All mutating will be group-specific.

```{r}
weather_df %>% 
  group_by(name) %>% 
  mutate(
    mean_tmax = mean(tmax, na.rm = TRUE),
    centered_tmax = tmax - mean_tmax
  ) %>% 
  ggplot(aes(x = date, y = centered_tmax, color = name)) +
  geom_point()
```

#### Window functions in grouped mutates
To get the rank in terms of order (first coldest day, 2nd, etc.), want to take all tmax values and sort and rank.

```{r}
# look at ?min_rank for description of windowed rank functions

weather_df %>% 
  group_by(name, month) %>% 
  mutate(
    tmax_rank = min_rank(tmax),
    tmax_rank_desc = min_rank(desc(tmax))
  ) %>% 
  arrange(name, month, tmax_rank) %>% 
  filter(tmax_rank == 1) # to get the coldest day in each location in each month
```

### lag

Look at change in max temperature from current day to next day

Opposite of the lag function is the lead function. 

```{r}
weather_df %>%
  group_by(name)  %>% 
  arrange(name, date) %>% 
  mutate(one_day_tmax_change = tmax - lag(tmax)) %>% summarize(sd_daily_chnge = sd(one_day_tmax_change, na.rm = TRUE))
```



